|

### 2022-12-23 paper.Rmd
Fri Dec 23 22:00:06 CET 2022: update-remote.sh

## Exploratory Phase

@@

It can also be observed that the population diversity, measured by the normalized levenshtein distance, is different between both DAE-GP variants. Both algorithms start at a unique rate of about 0.8 in the 0.th generation and lower rapidly to an average of about 0.45 in the 7.th generation. In the following generations, the diversity of pre-trained DAE-GP keeps lowering steadily while regular DAE-GP slowly increases the population diversity to a normalized levenshtein distance of about 0.55. This could be a first explanation for the better performance of regular DAE-GP in this experiment, by creating more diverse populations over the later generations of the search the algorithm is better at exploiting the search space than the pre-trained DAE-GP algorithm.

@@

To gather a deeper understanding about the effect of pre-training DAE-LSTM in DAE-GP a series of experiments was conducted, again using the airfoil dataset for symbolic regression, and different parameter configurations for the number of hidden layers as well as the number of hidden neurons per hidden layer.

@@

>>>>

<!---
## Exploratory Phase

@@

It can also be observed that the population diversity, measured by the normalized levenshtein distance, is different between both DAE-GP variants. Both algorithms start at a unique rate of about 0.8 in the 0.th generation and lower rapidly to an average of about 0.45 in the 7.th generation. In the following generations, the diversity of pre-trained DAE-GP keeps lowering steadily while regular DAE-GP slowly increases the population diversity to a normalized levenshtein distance of about 0.55. This could be a first explanation for the better performance of regular DAE-GP in this experiment, by creating more diverse populations over the later generations of the search the algorithm is better at exploiting the search space than the pre-trained DAE-GP algorithm.
-->

@@

To gather a deeper understanding about the effect of pre-training DAE-LSTM in DAE-GP a series of experiments was conducted using the airfoil dataset for symbolic regression while using and different parameter configurations for the number of hidden layers as well as the number of hidden neurons per hidden layer.

@@

## Effect on search performance

After closely examining the effect of pre-training on DAE-GPs generalization behaviour another series of experiments is conducted to study how pre-training influences the overall search behaviour of DAE-GP. Again the Airfoil problem was selected as a first benchmark problem to test pre-trained DAE-GP against traditional DAE-GP [@dae-gp_2022_symreg], the parameters are described in table $A$

```{r airfoil_fullRun_2hl_maxIndSize_params, echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
fig.pos="H"
knitr::kable(
  read.csv("./tables/airfoil_fullRun_2hl_maxIndSize_params"),
  caption="Airfoil - Parameter for full Run"
)%>% row_spec(0,bold=TRUE ) %>% kableExtra::kable_styling(latex_options = "hold_position")
```

Regarding the hidden neurons, the dimension of each hidden layer, I used two different strategies: Regular DAE-GP uses the same strategy as described by earlier work (see [@dae-gp_2022_symreg] or [@dae-gp_2020_rtree]), the number of hidden neurons is dynamically set at each generation to the maximum length of all solutions inside the current training population. For pre-trained DAE-GP another strategy had to be used, we use a fixed number of hidden neurons for each run that is set equal to the maximum length of all individuals that are present in the initial pre-training population. I decided to take this aproach instead of using a fixed number of hidden neurons for the fact that it simplifies the adjustment of the parameter for later experiments on different problems. It should nonetheless be mentioned that this strategy does give pre-training on average more hidden neurons per layer resulting in a more complex network (since the pre-training population is much larger than the regular population, the chances for larger individuals inside it are also higher).

![Airfoil - Best Fitness over 30 Generations](./img/airfoil_FullRun_30gens/mean_median_fitness_byGens.png)

![Airfoil - Fitness Distribution Boxplot](./img/airfoil_FullRun_30gens/final_fit_boxplot.png)

![Airfoil - Average Population Diversity](./img/airfoil_FullRun_30gens/mean_popDiversity_byGens.png)
Since @dae-gp_2022_symreg finds, that DAE-GP finds higher quality and smaller sized solutions smaller than average GP for a given number of 10.000 the airfoil problem

![Airfoil - Average Solution Size](./img/airfoil_FullRun_30gens/mean_Size_byGens.png)

### 2022-12-23 paper.tex
Fri Dec 23 22:00:06 CET 2022: update-remote.sh

Johannes Gutenberg University Mainz}
\author{}
\date{\vspace{-2.5em}Date of Submission: 2022-12-19}

@@

\hypertarget{exploratory-phase}{%
\subsection{Exploratory Phase}\label{exploratory-phase}}

In a first exploratory experiment DAE-GP was tested on the Aifoil
dataset to perform a symbolic regression task. The aim of this initial
experiment is to search for differences in behavior between an algorithm
that uses a pre-training strategy and regular DAE-GP. The parameters for
the experiment are summarized in table 2, the average results for
fitness, number of fitness evaluations, average tree size of the
population and the unique rate of the population are summarized in plot
1.

\caption{\label{tab:unnamed-chunk-1}Airfoil - Configuration}
\centering
\begin{tabular}[t]{l|l}
\hline
\textbf{Hyperparameter} & \textbf{Value}\\
\hline
Training/Testing Split & 50:50\\
\hline
Pre-Training Population Size & 10000\\
\hline
Number of Hidden Layers & 1\\
\hline
Selection & Tournament\\
\hline
Tournament Size & 7\\
\hline
\end{tabular}
\end{table}

\begin{figure}
\centering
\includegraphics{./img/airfoil_exploratory/airfoil_exploratory.png}
\caption{Airfoil - First Results}
\end{figure}

In this first experiment, the average quality of solutions produced by
the pre-trained algorithm shows to be lower than that produced by
regular DAE-GP. Differences in fitness start to show at generation 5
where regular DAE-GP starts to reduces RMSE more efficiently than the
pre-trained version. The gap in average fitness grows larger until the
end of the evolutionary search between both algorithms. Observing the
gap between RMSE on the training dataset and the testing dataset shows
no major differences, the training error is slightly lower than the
testing error for both algorithms and indicates a similar generalization
behavior.

Another interesting metric to investigate is the population's average
tree size, in general smaller solutions are preferable to their larger
counterparts because of improved interpretability. The results show,
that average tree size is reduced strongly in generations 1-5 for both
algorithms and than slowly increases back. Differences between regular
DAE-GP and the pre-trained DAE-GP start to show at generation 2, where
the average population tree size is continuously lower for regular
DAE-GP.

It can also be observed that the population diversity, measured by the
normalized levenshtein distance, is different between both DAE-GP
variants. Both algorithms start at a unique rate of about 0.8 in the
0.th generation and lower rapidly to an average of about 0.45 in the
7.th generation. In the following generations, the diversity of
pre-trained DAE-GP keeps lowering steadily while regular DAE-GP slowly
increases the population diversity to a normalized levenshtein distance
of about 0.55. This could be a first explanation for the better
performance of regular DAE-GP in this experiment, by creating more
diverse populations over the later generations of the search the
algorithm is better at exploiting the search space than the pre-trained
DAE-GP algorithm.

@@

>>>>

Johannes Gutenberg University Mainz}
\author{}
\date{\vspace{-2.5em}Date of Submission: 2022-12-23}

@@

@@

\hypertarget{effect-on-search-performance}{%
\subsection{Effect on search
performance}\label{effect-on-search-performance}}

After closely examining the effect of pre-training on DAE-GPs
generalization behaviour another series of experiments is conducted to
study how pre-training influences the overall search behaviour of
DAE-GP. Again the Airfoil problem was selected as a first benchmark
problem to test pre-trained DAE-GP against traditional DAE-GP
{[}\textbf{gp\_2022\_symreg?}{]}, the parameters are described in table
\(A\)

\caption{\label{tab:airfoil_fullRun_2hl_maxIndSize_params}Airfoil - Parameter for full Run}
\centering
\begin{tabular}[t]{l|l}
\hline
\textbf{parameter} & \textbf{value}\\
\hline
populationSize & 500\\
\hline
generations & 30\\
\hline
fitness & RMSE\\
\hline
TrainingMode & Convergence\\
\hline
SamplingSteps & 2\\
\hline
hiddenLayers & 2\\
\hline
Selection & Binary Tournament Selection\\
\hline
Pre-Training PopulationSize(Training/Validation) & 10000/100000\\
\hline
Pre-Training TrainingMode & Early Stopping\\
\hline
\end{tabular}
\end{table}

Regarding the hidden neurons, the dimension of each hidden layer, I used
two different strategies: Regular DAE-GP uses the same strategy as
described by earlier work (see {[}\textbf{gp\_2022\_symreg?}{]} or
{[}18{]}), the number of hidden neurons is dynamically set at each
generation to the maximum length of all solutions inside the current
training population. For pre-trained DAE-GP another strategy had to be
used, we use a fixed number of hidden neurons for each run that is set
equal to the maximum length of all individuals that are present in the
initial pre-training population.
